Human promoter directionality is determined by transcriptional initiation and the opposing activities of INTS11 and CDK9

Version 1

Abstract

Abstract

RNA polymerase II (RNAPII) transcription initiates bidirectionally at many human protein-coding genes. Sense transcription usually dominates and leads to messenger RNA production, whereas antisense transcription rapidly terminates. The basis for this directionality is not fully understood. Here, we show that sense transcriptional initiation is more efficient than in the antisense direction, which provides initial directionality. After transcription begins, the opposing functions of Integrator (INTS11) and cyclin-dependent kinase 9 (CDK9) maintain directionality. INTS11 terminates antisense transcription, whereas sense transcription is protected from INTS11-dependent attenuation by CDK9. INTS11 terminates transcription in both directions upon CDK9 inhibition, and the engineered recruitment of CDK9 prevents attenuation by INTS11. Therefore, transcriptional initiation and the opposing activities of CDK9 and INTS11 explain the attenuation of antisense transcription, the more extensive nature of sense transcription, and promoter directionality.

Introduction

In humans, most protein-coding genes initiate transcription bidirectionally. Usually, only sense transcription is extended and leads to mRNA synthesis. Antisense transcription terminates within ∼1 kilobase (kb), and the short non-coding (nc)RNA is degraded1. Similar asymmetry frequently occurs in unicellular eukaryotes (e.g., budding yeast) and at some plant promoters2, 3. Thus, promoter directionality is a broadly observed phenomenon. Interestingly, bidirectionality is the ground state of promoters and directionality is acquired over evolutionary time4. This is proposed to be through a combination of DNA sequences and proteins that favour the directional initiation and elongation of transcription.

The present explanation for the directionality of mammalian RNAPII promoters involves the arrangement of U1 snRNA binding sites and polyadenylation signals (PASs). U1 snRNA promotes RNAPII elongation through protein-coding genes by binding to RNA and preventing early termination5–7. It does this by silencing PASs and antagonizing other attenuation mechanisms, which include the PP1 regulator, PNUTS, and the restrictor complex8, 9. In contrast, U1 binding splice sites are rarer in short antisense transcripts, which are often rich in PAS sequences that are proposed to promote transcriptional termination10. This model predicts that polyadenylation factors control a large fraction of antisense transcriptional termination.

The multi-subunit Integrator complex also regulates promoter-proximal transcription11–15. The Integrator complex comprises the backbone, arm, phosphatase, and endonuclease modules11. Its endoribonuclease is INTS11, and its phosphatase activity is mediated by INTS6 and protein phosphatase 2A (PP2A). INTS11 endonuclease broadly affects promoter-proximal transcriptional attenuation whereas INTS phosphatase is proposed to regulate the escape of RNAPII into elongation16, 17. INTS6/PP2A phosphatase functionally antagonises CDK9, which is vital for RNAPII promoter escape and productive elongation18. By this model, CDK9 activity promotes elongation across protein-coding genes and INTS6/PP2A opposes it.

We tested the prediction that PAS factors control transcription directionality by terminating antisense transcription, but this is not usually the case. Instead, we find that promoter directionality is conferred by preferential initiation in the sense direction and is thereafter maintained by INTS11 and CDK9. The termination of most antisense transcription is constitutively INTS11-dependent, whereas sense transcription is hypersensitive to INTS11 only when CDK9 activity is simultaneously inhibited. We hypothesise that CDK9 activity desensitises sense transcription to INTS11 and that reduced CDK9 activity in the antisense direction exposes RNAPII to INTS11-dependent termination.

Results and discussion

The present explanation of mammalian promoter directionality invokes early PAS-dependent termination of antisense transcription10. This is partly based on the direct detection of polyadenylated antisense transcripts, which provides evidence that some transcriptional termination is PAS-dependent. However, there are non-polyadenylated antisense RNAs that might be attenuated in other ways19. To test the relevance of PAS-dependent termination for antisense transcriptional termination and promoter directionality we tagged RBBP6 with the dTAG degron. Because RBBP6 is required to activate the PAS cleavage machinery20, 21, its depletion should inhibit any PAS-dependent transcriptional termination. Three homozygous dTAG-RBBP6 clones were isolated (Supplemental Figure 1A) and tagged RBBP6 was efficiently depleted after exposure to the dTAGv-1 degrader (Figure 1A). To test the overall contribution of RBBP6 to transcription, we used POINT (Polymerase Intact Nascent Transcript)-seq22, which maps nascent transcription by sequencing full-length RNA extracted from immunoprecipitated RNAPII.

RBBP6 loss disrupts PAS-dependent termination of sense transcription.

A. Western blot demonstrating the depletion of dTAG-RBBP6 over a time course of dTAGv-1 addition. SPT5 serves as a loading control.

B. Genome browser track of NEDD1 in POINT-seq data from dTAG-RBBP6 cells treated or not (2hr) with dTAGv-1. RBBP6 depletion induces a termination defect in the protein-coding direction (downstream of the indicated PAS) but not the upstream antisense direction. The Y-axis shows Reads Per Kilobase per Million mapped reads (RPKM).

C. Metaplot of POINT-seq data from dTAG-RBBP6 cells treated (RBBP6 dep) or not (Ctrl) (2hr) with dTAGv-1. This shows 1316 protein-coding genes selected as separated from any expressed transcription unit by ≥10kb. Signals above and below the x-axis are sense and antisense reads, respectively. The Y-axis scale is RPKM. TSS=transcription start site; TES=transcription end site (this marks the PAS position).

D. Heatmap representation of the data in C, which displays signal as a log2 fold change (log2FC) in RBBP6 depleted versus un-depleted conditions.

Figure 1B shows POINT-seq coverage over NEDD1 including the upstream antisense transcript. RBBP6 loss causes a termination defect beyond the PAS shown by the extended POINT-seq signal beyond the NEDD1 gene. However, RBBP6 does not affect the termination of upstream antisense RNA which is not extended when RBBP6 is depleted. Metaplot and heatmap analyses across 1316 genes confirm this as a global trend (Figures 1C and D). Although many antisense transcripts contain multiple AAUAAA sequences, most still terminate RBBP6-independently (Supplementary Figures 1B and C). Similarly, we recently showed that the PAS-dependent 5’→3’ exonucleolytic torpedo terminator, XRN2, does not affect antisense transcriptional termination9. Therefore, although a fraction of antisense transcription is polyadenylated, most of it can terminate using PAS-independent mechanisms.

Our data argue that PAS-independent termination mechanisms are prevalent for antisense transcription. A major PAS-independent termination pathway is driven by the Integrator complex, which was first identified as the 3’ end processing complex for snRNAs23. Several reports show that Integrator terminates transcription from most promoters, including those that initiate antisense transcription, so might affect promoter directionality11, 16, 17. To analyse this, we tagged its endonucleolytic subunit (INTS11) with a dTAG degron24, which enables rapid depletion of its catalytic activity (Figure 2A). We then performed POINT-seq on INTS11-dTAG cells depleted or not of INTS11 to assay the global impact of INTS11 on RNAPII transcription. The established effect of INTS11 at snRNAs was detected in our POINT-seq data and demonstrates the efficacy of this approach (Figure 2B).

INTS11 loss disrupts the termination of antisense transcription.

A. Western blot demonstrating homozygous tagging of INTS11 with dTAG and the depletion of INTS11-dTAG after 1.5 hr treatment with dTAGv-1. INTS8 serves as a loading control.

B. Genome browser track showing POINT-seq signal over RNU5A-1 and RNU5B-1 in INTS11-dTAG cells treated or not (1.5 hr) with dTAGv-1. Note, that INTS11 depletion induces a termination defect in each case. Y-axis shows RPKM following spike-in normalisation.

C. Genome browser track showing POINT-seq signal over PGK1 and its upstream antisense region in INTS11-dTAG cells treated or not (1.5 hr) with dTAGv-1. Y-axis shows RPKM following spike-in normalisation.

D. Metaplot of POINT-seq data from INTS11-dTAG cells treated or not (1.5 hr) with dTAGv-1. This shows 1316 expressed protein-coding genes that are separated from any expressed transcription unit by ≥10kb. Signals above and below the x-axis are sense and antisense reads, respectively. Y-axis shows RPKM following spike-in normalisation.

E. Heatmap representation of the data in D, which displays signal as a log2 fold change (log2FC) in INTS11 depleted versus undepleted conditions over a region 3kb upstream and downstream of annotated TSSs.

As exemplified by PGK1 (Figure 2C), INTS11 does not affect the termination of protein-coding transcription but its loss causes a strong upregulation of antisense transcription. Meta-analysis shows the generality of antisense transcriptional attenuation via INTS11 (Figure 2D) and heatmap analysis demonstrates the dominant impact of INTS11 on antisense versus sense transcription at most promoters (Figure 2E). Transcription over protein-coding gene bodies is generally more modestly affected by INTS11 loss, but there is evidence of mildly increased transcription over the 5’ end of some genes as was recently reported (Supplemental Figure 2A)12–17. Overall, INTS11 frequently attenuates antisense transcription whereas a smaller fraction of sense transcription is affected.

Our analyses show that antisense RNAs are more frequently terminated via INTS11 than sense transcripts, which ultimately terminate via the PAS-dependent mechanism. The sensitivity of antisense transcripts to INTS11 might be due to the nature of antisense transcription or some other promoter feature. To interrogate this further, we analysed two other promoter classes: those where protein-coding transcripts are initiated in both directions and those that initiate the bidirectional transcription of unstable enhancer (e)RNAs. In the former case, both transcripts are extended and stable (i.e., like most sense transcripts); in the latter case, both are short and unstable like most antisense transcripts. When both directions are protein-coding, INTS11 depletion causes modest reductions in transcription (Supplemental Figure 2B). Conversely, eRNA transcription was bidirectionally upregulated upon INTS11 elimination like the antisense transcripts in (Supplemental Figure 2C). We conclude that short ncRNAs are more strongly affected by INTS11 than protein-coding transcripts. At directional promoters this results in the attenuation of antisense transcription.

The increased antisense POINT-seq signal following INTS11 loss is consistent with defects in transcriptional termination. However, it could result from increased transcriptional initiation. We were also interested in whether preferential sense initiation could contribute to observed directionality as was proposed in budding yeast4. To precisely resolve directional aspects of initiation and assay any impact of INTS11, we devised a variant of POINT-seq called short (s)POINT. Briefly, sPOINT follows the POINT-seq protocol but library preparation employs the selective amplification of 5’ capped RNAs <150nts (Figure 3A, Supplemental Figure 3A, Experimental Procedures). The sPOINT signal over PGK1 exemplifies this and demonstrates a very restricted signal close to the TSS (Figure 3B). Figure 3C compares the meta profile of POINT- and sPOINT-seq. While POINT-seq reads span entire transcribed sequences the sPOINT signal is restricted to the promoter-proximal region. Thus, the well-known promoter-proximal RNAPII pause is efficiently assayed by sPOINT-seq. Our metaplot in Figure 3C shows that most capped RNAPII-associated RNA <150nts are promoter-proximal and thus relate to paused polymerases.

Transcription initiation is more efficient and focused in the sense direction.

A. Schematic of sPOINT-seq protocol. The POINT-seq protocol is followed, in which chromatin is isolated and engaged RNAPII is immunoprecipitated. Short transcripts are preferentially amplified during library preparation (see Experimental Procedures for full details).

B. Comparison of POINT-(top trace) and sPOINT-seq (lower trace) on PGK1. Y-axis units are RPKM.

C. Metaplot comparison of POINT-(top plot) and sPOINT-seq (lower plot) profiles across the 684 highest expressed protein-coding that are separated from expressed transcription units by ≥10kb.. Signals above and below the x-axis are sense and antisense reads, respectively. Y-axis shows RPKM following spike-in normalisation.

D. Top metaplot shows full read coverage for sPOINT-seq performed in INTS11-dTAG cells treated or not (1.5hr) with dTAGv-1. The lower metaplot is the same data but only the 5’ end of each read is plotted. Both plots display a region +/-1kb from the TSS of the top 20% (based on sPOINT signal) promoters. Y-axis signals are RPKM following spike-in normalisation.

E. Metaplot zoom of the antisense TSS signals deriving from the lower plot in D. This makes clear the dispersed sites of initiation. The Y-axis scale is RPKM following spike-in normalisation.

F. Genome browser track of PGK1 promoter region in sPOINT-seq. This showcases the focused sense TSS (black arrows) and the dispersed antisense reads (brackets) generalised in the metaplots in D and E. Note the higher y-axis scale (RPKM) for sense vs. antisense.

We then performed sPOINT-seq in INTS11-dTAG cells treated or not with dTAGv-1 and plotted the coverage of full-length reads within 1kb of TSSs (Figure 3D). Because sPOINT-seq maps the 5’ and 3’ ends of these reads, it also detects TSSs at single-nucleotide resolution and these 5’ ends are plotted in the lower meta profile in Figure 3D. INTS11 loss results in lower sPOINT-seq signal arguing against increased transcriptional initiation in its absence. Together with the increased POINT-seq – especially for antisense transcription – this indicates more RNAPII promoter escape upon INTS11 loss. Interestingly, this experiment reveals sense transcriptional initiation to be more efficient and focused compared to the antisense direction. This is clear from the TSS-aligned metaplot in Figure 3D and the dispersed nature of antisense transcription shown in Figure 3E. Quantitation of the TSS-derived sense vs. antisense reads demonstrates the higher signal in the sense direction in untreated cells (Supplemental Figure 3B). Figure 3F (PGK1) and supplemental figure 3C (ACTB) exemplify these features on individual genes. INTS11 loss does not dramatically alter TSS position or focus. Although a lower resolution technique, RNAPII chromatin-immunoprecipitation and sequencing (ChIP-seq) confirmed these promoter characteristics and the mild impact of INTS11 (Supplemental Figures 3D and E). Overall, these data show that directionality is partly determined by preferential initiation in the sense direction.

After initiation, directionality is maintained because it is well-established that sense transcription goes further than antisense transcription. If INTS11 regulates this, an opposing force is required to prevent it from attenuating sense transcription. INTS11 affects transcription very early, occupies promoters, and becomes less active as RNAPII moves into elongation16–18, 25, 26. Therefore, if INTS11 is counteracted to allow sense transcription, any responsible mechanism needs to act early. One of the first transcriptional checkpoints involves the phosphorylation of RNAPII and other factors by CDK9, which releases promoter-proximally paused RNAPII into elongation27. During this process, the Integrator-associated phosphatase, PP2A, antagonizes CDK9 and presumably regulates the sensitivity of RNAPII to INTS1118. Because Integrator phosphatase has little effect on antisense transcription16, we hypothesised that INTS11 sensitivity and CDK9 activity are inversely correlated to maintain directionality after initiation.

This hypothesis predicts that sense transcription will be attenuated via INTS11 when CDK9 is inactive. To test this genome-wide, we depleted INTS11 from INTS11-dTAG cells in the presence or absence of a specific CDK9 inhibitor (NVP-228) and performed POINT-seq. As exemplified by CDCA7, the depletion of INTS11 alone caused an antisense transcriptional termination defect with a milder impact on the protein-coding sense direction (Figure 4A). As expected, NVP-2 treatment reduced transcription over the protein-coding gene body. Antisense transcription also displays some CDK9 sensitivity. Importantly, in NVP-2-treated cells, INTS11 loss increased transcription in both directions - effectively compromising the maintenance of directionality. Further examples are displayed in Supplemental Figures 4A and B. This is a genome-wide trend as shown in the metaplots in Figures 4B and C. Heatmap analysis of the effect of INTS11 loss after CDK9 inhibition shows bidirectional upregulation of transcription at most assayed genes (Figure 4D). This experiment shows a clearer effect of INTS11 loss over the 5’ end of sense transcripts versus our experiment in Figure 2 (although this is again more modest than the antisense effect). We speculate this to be due to the slightly extended period of INTS11 depletion employed alongside NVP-2 treatment (2.5 hr here vs. 1.5 hr for Figure 2).

Promoters lose their directionality when CDK9 is inhibited.

A. Genome browser track of CDCA7 in POINT-seq data derived from INTS11-dTAG cells either untreated, dTAG treated, NVP-2 treated or dTAG and NVP-2 treated (2.5 hr). Signals above and below the Y-axis are sense and antisense reads, respectively. The Y-axis scale shows RPKM following spike-in normalisation.

B. Metaplot of POINT-seq analysis of POINT-seq in INTS11-dTAG cells depleted or not of INTS11 and treated or not with NVP-2 (2.5 hr). This shows 1316 protein-coding genes selected as separated from any expressed transcription unit by ≥10kb. The regions 3kb upstream and downstream of genes are included. Y-axis units are RPKM following spike-in normalisation.

C. Metaplot of the same CDK9i + and -dTAG data shown in B but zoomed into the region 3kb upstream and downstream of the TSS.

D. Heatmap representation of the data in C, which displays signal as a log2 fold change (log2FC) in INTS11 depleted versus un-depleted conditions covering a region 3kb upstream and downstream of the TSS.

E. qRT-PCR analysis of INTS11-dTAG cells transfected with the HIV reporter construct with or without TAT then depleted or not of INTS11. Quantitation shows signals relative to those obtained in the presence of INTS11 and the absence of TAT after normalising to MALAT1 RNA. n=3. Error bars show standard deviation. ** denoted p=0.01. Note that INTS11 depletion was performed concurrently with transection (14hr in total).

F. Model for promoter directionality depicting higher levels of focused transcriptional initiation in the sense direction together with opposing gradients of CDK9 and INTS11 activity that peak in sense and antisense directions, respectively.

Following CDK9 inhibition and INTS11 loss, the largest recovery of protein-coding transcription is often over the 5’ end of genes. This presumably reflects a continued requirement of CDK9 for elongation, even when INTS11 is absent. Furthermore, Integrator phosphatase remains RNAPII-associated after acute INTS11 loss17, which would antagonise elongation especially if CDK9 is lacking. Importantly, the POINT-seq signal remains higher in the sense vs. antisense direction even when CDK9 and INTS11 are both compromised. This is consistent with our sPOINT-seq conclusion that levels of sense transcriptional initiation are higher compared to the antisense direction. Finally, we confirmed the CDK9-mediated suppression of INTS11 on four selected protein-coding genes using an alternative inhibitor, 5, 6-dichloro-1-β-D-ribofuranosylbenzimidazole (DRB) (Supplemental Figure 4C). In sum, most RNAPII promoters are bidirectionally affected by INTS11 loss when CDK9 is inhibited.

If CDK9 counteracts INTS11, its recruitment should also prevent transcriptional attenuation by Integrator. To assay this, we employed a plasmid where transcription is driven by the human immunodeficiency virus (HIV) promoter. Transcription from the HIV promoter results in the synthesis of the trans-activating response (TAR) element and promoter-proximal RNAPII pausing. Pause release requires the trans-activator of transcription (TAT), which promotes RNAPII elongation by recruiting CDK929. INTS11 suppresses transcription from the HIV promoter when TAT is not present30. To test the effect of CDK9 on this process, INTS11-dTAG cells were transfected with the HIV reporter with or without TAT before treatment or not with dTAGv-1. Transcription from the reporter was analysed by qRT-PCR (Figure 4E). HIV transcription is induced by INTS11 loss in the absence of TAT. Transcription was greatly stimulated by TAT (∼200 fold) but was no longer enhanced by INTS11 loss (there was slightly less reporter RNA recovered). Therefore, CDK9 recruitment by TAT alleviates INTS11-dependent attenuation of transcription.

How CDK9 opposes INTS11 is unresolved but, of possible relevance, INTS11 and SPT5 are adjacent in the RNAPII: Integrator structure31. As SPT5 is a substrate of CDK932–34, its phosphorylation might evict INTS11 or prevent its association with the complex. We show that sense transcriptional initiation dominates over the antisense direction. This could reflect the positioning of DNA elements that attract initiation factors to advantage sense transcription4. Consistently, mammalian initiation components show peak occupancy over annotated sense TSSs35. Thereafter, the maintenance of directionality by CDK9 and INTS11 is consistent with their well-established roles in promoting and terminating transcription. Published ChIP-seq indicates their respective enrichment in sense and antisense directions, which supports our model (Supplemental Figure 4D18). Nevertheless, some antisense transcription is NVP-2 sensitive, revealing CDK9 activity on a fraction of RNAPII (Figure 4). As CDK9 opposes INTS11, this fraction of antisense transcription may be terminated via other mechanisms like restrictor and the PAS-dependent machinery9, 10, 36–38. Overall, transcriptional initiation and the opposing activities of CDK9 and INTS11 establish and maintain promoter directionality (see model in Figure 4F).

eLife Assessment

The important study uses a new experimental method to provide compelling evidence on how sense- and anti-sense transcription is differentially regulated. The method described here can generally be used to study the alterations in transcription. This paper will be of interest to scientists working in the gene regulation community.

Reviewer #1 (Public Review):

Summary:

In this manuscript, Eaton et al. examine the regulation of transcription directionality using a powerful genomic approach (more about the methodology below). Their data challenge the notion that the polyadenylation signal-reading Cleavage and Polyadenylation (CPA) complex is responsible for controlling promoter directionality by terminating antisense transcription. Namely, depletion of the required CPA factor RBBP6 has little effect on antisense transcription measured by POINT. They find instead that initiation is intrinsically preferential in the sense direction and additionally maintained by the activities of an alternative processing complex called Integrator, together with the kinase CDK9. In the presence of CDK9 activity, depletion of Integrator endoribonuclease INTS11 leads to globally increased transcription in the antisense direction, and minor effects in the sense direction. However, CDK9 inhibition reveals that sense transcription is also sensitive to INS11 depletion. The authors suggest that CDK9 activity is stronger in the sense direction, preventing INTS11-mediated premature termination of sense transcrpts.

Strengths:

The combination of acute depletion of the studied factors using degron approaches (important to limit possible secondary effects), together with novel and very sensitive nascent transcriptomics methods POINT and sPOINT is very powerful. The applied spike-in normalization means the analysis is more rigorous than most. Using this methodology allowed the authors to revisit the interesting question of how promoter/transcription directionality is determined.

The data quality appears very good and the fact that both global analysis as well as numerous gene-specific examples are shown makes it convincing.

The manuscript is well written and hence a pleasure to read.

Weaknesses:

I am slightly worried about the reproducibility of the data - it is unclear to me from the manuscript if and which experiments were performed in replicate (lack of table with genomic experiments and GEO access, mentioned in more detail in below recommendations to authors), and the methods could be more detailed.

A separate discussion section would be useful, particularly since the data provided challenge some concepts in the field. How do the authors interpret U1 data from the Dreyfuss lab in light of their results? How about the known PAS-density directionality bias (more PAS present in antisense direction than in sense) - could the differential PAS density be still relevant to transcription directionality?

I find that the provided evidence for promoter directionality to be for the most part due to preferential initiation in the sense direction should be stressed more. This is in my eyes the strongest effect and is somehow brushed under the rug.

References 12-17 report an effect of Integrator on 5' of protein-coding genes, while data in Figure 2 appears contradictory. Then, experiments in Figure 4 show a global effect of INST11 depletion on promoter-proximal sense transcription. In my opinion, data from the 2.5h time-point of depletion should be shown alongside 1.5h in Figure 2 so that it is clear that the authors found an effect similar to the above references. I find the current presentation somehow misleading.

Conclusion/assessment:

This important work substantially advances our understanding of the mechanisms governing the directionality of human promoters. The evidence supporting the claims of the authors is compelling, with among others the use of advanced nascent transcriptomics including spike-in normalization controls and acute protein depletion using degron approaches.

In my opinion, the authors' conclusions are in general well supported.

Not only the manuscript but also the data generated will be useful to the wide community of researchers studying transcriptional regulation. Also, the POINT-derived novel sPOINT method described here is very valuable and can positively impact work in the field.

Reviewer #2 (Public Review):

Summary:

Eaton and colleagues use targeted protein degradation coupled with nascent transcription mapping to highlight a role for the integrator component INST11 in terminating antisense transcription. They find that upon inhibition of CDK9, INST11 can terminate both antisense and sense transcription - leading to a model whereby INST11 can terminate antisense transcription and the activity of CDK9 protects sense transcription from INST11-mediated termination. They further develop a new method called sPOINT which selectively amplifies nascent 5' capped RNAs and find that transcription initiation is more efficient in the sense direction than in the antisense direction. This is an excellent paper that uses elegant experimental design and innovative technologies to uncover a novel regulatory step in the control of transcriptional directionality.

Strengths:

One of the major strengths of this work is that the authors endogenously tag two of their proteins of interest - RBBP6 and INST11. This tag allows them to rapidly degrade these proteins - increasing the likelihood that any effects they see are primary effects of protein depletion rather than secondary effects. Another strength of this work is that the authors immunoprecipitate RNAPII and sequence extracted full-length RNA (POINT-seq) allowing them to map nascent transcription. A technical advance from this work is the development of sPOINT which allows the selective amplification of 5' capped RNAs < 150 nucleotides, allowing the direction of transcription initiation to be resolved.

Weaknesses:

While the authors provide strong evidence that INST11 and CDK9 play important roles in determining promoter directionality, their data suggests that when INST11 is degraded and CDK9 is inhibited there remains a bias in favour of sense transcription (Figures 4B and C). This suggests that there are other unknown factors that promote sense transcription over antisense transcription and future work could look to identify these.

Reviewer #3 (Public Review):

Summary:

Using a protein degradation approach, Eaton et al show that INST11 can terminate the sense and anti-sense transcription but higher activity of CDK9 in the sense direction protects it from INS11-dependent termination. They developed sPOINT-seq that detects nascent 5'-capped RNA. The technique allowed them to reveal robust transcription initiation of sense-RNA as compared to anti-sense.

Strengths:

The strength of the paper is the acute degradation of proteins, eliminating the off-target effects. Further, the paper uses elegant approaches such as POINT and sPOINT-seq to measure nascent RNA and 5'-capped short RNA. Together, the combination of these three allowed the authors to make clean interpretations of data.

Weaknesses:

While the manuscript is well written, the details on the panel are not sufficient. The methods could be elaborated to aid understanding. Additional discussion on howthe authors' findings contradict the existing model of anti-sense transcription termination should be added.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

In this manuscript, Eaton et al. examine the regulation of transcription directionality using a powerful genomic approach (more about the methodology below). Their data challenge the notion that the polyadenylation signal-reading Cleavage and Polyadenylation (CPA) complex is responsible for controlling promoter directionality by terminating antisense transcription. Namely, depletion of the required CPA factor RBBP6 has little effect on antisense transcription measured by POINT. They find instead that initiation is intrinsically preferential in the sense direction and additionally maintained by the activities of an alternative processing complex called Integrator, together with the kinase CDK9. In the presence of CDK9 activity, depletion of Integrator endoribonuclease INTS11 leads to globally increased transcription in the antisense direction, and minor effects in the sense direction. However, CDK9 inhibition reveals that sense transcription is also sensitive to INS11 depletion. The authors suggest that CDK9 activity is stronger in the sense direction, preventing INTS11-mediated premature termination of sense transcrpts.

Strengths:

The combination of acute depletion of the studied factors using degron approaches (important to limit possible secondary effects), together with novel and very sensitive nascent transcriptomics methods POINT and sPOINT is very powerful. The applied spike-in normalization means the analysis is more rigorous than most. Using this methodology allowed the authors to revisit the interesting question of how promoter/transcription directionality is determined.

The data quality appears very good and the fact that both global analysis as well as numerous gene-specific examples are shown makes it convincing.

The manuscript is well written and hence a pleasure to read.

We appreciate this positive assessment.

Weaknesses:

I am slightly worried about the reproducibility of the data - it is unclear to me from the manuscript if and which experiments were performed in replicate (lack of table with genomic experiments and GEO access, mentioned in more detail in below recommendations to authors), and the methods could be more detailed.

All sequencing data was deposited with GEO. Multiple biological replicates were performed for each sequencing experiment. Bigwig files are presented as a table in the GEO submissions. This data has now been made public.

A separate discussion section would be useful, particularly since the data provided challenge some concepts in the field. How do the authors interpret U1 data from the Dreyfuss lab in light of their results? How about the known PAS-density directionality bias (more PAS present in antisense direction than in sense) - could the differential PAS density be still relevant to transcription directionality?

As suggested, we have expanded our discussion to relate our findings to existing data. We think the results from the Dreyfuss lab are very important and highlight the role of U1 snRNA in enforcing transcriptional elongation. It does this in part by shielding PAS sequences. Recent work from our lab also shows that U1 snRNA opposes the Restrictor complex and PNUTS, which otherwise suppress transcription (Estell et al., Mol Cell 2023). Most recently, the Adelman lab has demonstrated that U1 snRNA generally enhances transcription elongation (Mimoso and Adelman., Mol Cell 2023). Our work does not challenge and is not inconsistent with these studies.

The role of U1 in opposing PAS-dependent termination inspired the idea that antisense transcriptional termination may utilise PASs. This was because such regions are rich in AAUAAA and comparatively poor in U1 binding sites. However, our RBBP6 depletion and POINT-seq data suggest that PAS-dependent termination is uncommon in the antisense direction. As such, other mechanisms suppress antisense transcription and influence promoter directionality. In our paper, we propose a major role for the Integrator complex.

We do not completely rule out antisense PAS activity and discuss the prior work that identified polyadenylated antisense transcripts. Nevertheless, this was detected by oligo-dT primed RT-PCR/Northern blotting, which cannot determine the fraction of non-polyadenylated RNA that could result from PAS-independent termination (e.g. by Integrator). To do that requires an analysis of total nascent transcription as achieved by our POINT-seq. Based on these experiments, Integrator depletion has a greater impact on antisense transcription than RBBP6 depletion.

I find that the provided evidence for promoter directionality to be for the most part due to preferential initiation in the sense direction should be stressed more. This is in my eyes the strongest effect and is somehow brushed under the rug.

We agree that this is an important finding and incorporated it into the title and abstract. As the reviewer recommends, we now highlight it further in the new discussion.

References 12-17 report an effect of Integrator on 5' of protein-coding genes, while data in Figure 2 appears contradictory. Then, experiments in Figure 4 show a global effect of INST11 depletion on promoter-proximal sense transcription. In my opinion, data from the 2.5h time-point of depletion should be shown alongside 1.5h in Figure 2 so that it is clear that the authors found an effect similar to the above references. I find the current presentation somehow misleading.

We are grateful for this suggestion and present new analyses demonstrating that our experiment in Figure 2 concurs with previous findings (Supplemental Figures 2A and B). Our original heatmap (Figure 2E) shows a very strong and general antisense effect of INTS11 loss. On the same scale, the effects in the sense direction are not as apparent, which is also the case using metaplots. New supplemental figure 2A now shows sense transcription from this experiment in isolation and on a lower scale, demonstrating that a subset of genes shows promoter-proximal increases in transcription following INTS11 depletion. This is smaller and less general than the antisense effect but consistent with previous findings. Indeed, our new analysis in supplemental figure 2B shows that affected protein-coding genes are lowly expressed, in line with Hu et al., Mol Cell 2023. This explains why a sense effect is not as apparent by metaplot, for which highly expressed genes contribute the most signal.

As a result of our analyses, we are confident that the apparently larger effect at the 2.5hr timepoint (Figure 4) that we initially reported is due to experimental variability and not greater effects of extended INTS11 depletion. Overlaying the 1.5h and 2.5h datasets (Supplemental Figure 4B) revealed a similar number of affected protein-coding genes with a strong (83%) overlap between the affected genes. To support this, we performed qPCR on four affected protein-coding transcripts which revealed no significant difference in the level of INTS11 effect after 2.5h vs 1.5h (Supplemental Figure 4C).

We now present data for merged replicates in Figures 2 and 4 which reveal very similar average profiles for -INTS11 vs +INTS11 at both timepoints. Overall, we believe that we have resolved this discrepancy by showing that it amounts to experimental variability and because the most acutely affected protein-coding genes are lowly expressed. As detailed above, we show this in multiple ways (and validate by qPCR) We have revised the text accordingly and removed our original speculation that differences reflected the timeframe of INTS11 loss.

Conclusion/assessment:

This important work substantially advances our understanding of the mechanisms governing the directionality of human promoters. The evidence supporting the claims of the authors is compelling, with among others the use of advanced nascent transcriptomics including spike-in normalization controls and acute protein depletion using degron approaches.

In my opinion, the authors' conclusions are in general well supported.

Not only the manuscript but also the data generated will be useful to the wide community of researchers studying transcriptional regulation. Also, the POINT-derived novel sPOINT method described here is very valuable and can positively impact work in the field.

We are grateful for the reviewers' positive assessment of our study.

Reviewer #2 (Public Review):

Summary:

Eaton and colleagues use targeted protein degradation coupled with nascent transcription mapping to highlight a role for the integrator component INST11 in terminating antisense transcription. They find that upon inhibition of CDK9, INST11 can terminate both antisense and sense transcription - leading to a model whereby INST11 can terminate antisense transcription and the activity of CDK9 protects sense transcription from INST11-mediated termination. They further develop a new method called sPOINT which selectively amplifies nascent 5' capped RNAs and find that transcription initiation is more efficient in the sense direction than in the antisense direction. This is an excellent paper that uses elegant experimental design and innovative technologies to uncover a novel regulatory step in the control of transcriptional directionality.

Strengths:

One of the major strengths of this work is that the authors endogenously tag two of their proteins of interest - RBBP6 and INST11. This tag allows them to rapidly degrade these proteins - increasing the likelihood that any effects they see are primary effects of protein depletion rather than secondary effects. Another strength of this work is that the authors immunoprecipitate RNAPII and sequence extracted full-length RNA (POINT-seq) allowing them to map nascent transcription. A technical advance from this work is the development of sPOINT which allows the selective amplification of 5' capped RNAs < 150 nucleotides, allowing the direction of transcription initiation to be resolved.

We appreciate this positive assessment.

Weaknesses:

While the authors provide strong evidence that INST11 and CDK9 play important roles in determining promoter directionality, their data suggests that when INST11 is degraded and CDK9 is inhibited there remains a bias in favour of sense transcription (Figures 4B and C). This suggests that there are other unknown factors that promote sense transcription over antisense transcription and future work could look to identify these.

We agree that other (so far, unknown) factors promote sense transcription over antisense, which was demonstrated by our short POINT. We have provided an expanded discussion on this in the revision. In our opinion, demonstrating that sense transcription is driven by preferential initiation in that direction is a key finding and we agree that the identification of the underlying mechanism constitutes an interesting avenue for future study.

Reviewer #3 (Public Review):

Summary:

Using a protein degradation approach, Eaton et al show that INST11 can terminate the sense and anti-sense transcription but higher activity of CDK9 in the sense direction protects it from INS11-dependent termination. They developed sPOINT-seq that detects nascent 5'-capped RNA. The technique allowed them to reveal robust transcription initiation of sense-RNA as compared to anti-sense.

Strengths:

The strength of the paper is the acute degradation of proteins, eliminating the off-target effects. Further, the paper uses elegant approaches such as POINT and sPOINT-seq to measure nascent RNA and 5'-capped short RNA. Together, the combination of these three allowed the authors to make clean interpretations of data.

We appreciate this positive assessment.

Weaknesses:

While the manuscript is well written, the details on the panel are not sufficient. The methods could be elaborated to aid understanding. Additional discussion on how the authors' findings contradict the existing model of anti-sense transcription termination should be added.

We have added more detail to the figure panels, which we hope will help readers to navigate the paper more easily. Specifically, the assay employed for each experiment is indicated in each figure panel. As requested, we provide a new and separate discussion section in the revision.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Congratulations on this important piece of work!

Some specific suggestions.

MAJOR

-The data are not available (Accession "GSE243266" is currently private and is scheduled to be released on Sep 01, 2026.) This should be corrected and as a minimum, the raw sequencing files as well as the spike-in scaled bigwig files should be provided in GEO.

We have made the data public. Raw and bigwig files are provided as part of the GEO upload.

MINOR

- It would be useful for readers if you could include catalog numbers of the reagents used in the study.

We have included this information in our revision.

- A table in experimental procedures summarizing the genomic experiments performed in this study as well as published ones reanalyzed here would be helpful.

This is now provided as part of the resources table.

- It would be easier for reviewers to evaluate the manuscript if the figure legends were included together with the figures on one page. This is now allowed by most journals.

We have used this formatting in the revision.

- Providing some captions for the results sections would be helpful.

We have included subheadings as suggested.

Reviewer #2 (Recommendations For The Authors):

Generally, I would suggest writing the experiment-type above panels where it is not immediately obvious what they are so a reader can appreciate the figures without referencing the legend. E.g. write POINT-seq on Figure 1B just to make it obvious to someone looking at the figures what methodology they are looking at. Likewise, you could write RNAPII ChIP-seq for Supplementary Figures 3D and 3E.

We have carried out this recommendation.

Can a y-axis be indicated on POINT-seq genome browser tracks? This could make them easier to interpret.

Y-axis scales are provided as RPKM as stated in the figure legends.

The authors could address/speculate in the text why there is less POINT-seq signal for the antisense transcript in the treatment condition in Figure 1B? Or could consider including a different example locus where this is not the case for clarity.

Acute depletion of poly(A) factors (like RBBP6) results in a strong read-through beyond the poly(A) signal of protein-coding genes as Figure 1 shows. However, it also causes a reduction in transcription levels, which can be seen in the figure and is correctly noted by the reviewer in this comment. We see this with other poly(A) factor depletions (e.g. CPSF73 and CPSF30 – Eaton et al., 2020 and Estell et al., 2021) and other labs have observed this too (e.g for CPSF73-dTAG depletion (Cugusi et al., Mol Cell 2022)). Plausible reasons include a limited pool of free RNAPII due to impaired transcriptional termination or limited nucleotide availability due to their incorporation within long read-through transcripts. For these reasons, we have retained the example in Figure 1B as a typical representation of the effect. Moreover, the heatmap in Figure 1D fairly represents the spectrum of effects following RBBP6 loss – highlighting the strong read-through beyond poly(A) signals and the marginal antisense effects.

"The established effect of INTS11 at snRNAs was detected in our POINT-seq data and demonstrates the efficacy of this approach (Figure 2B)." The authors could explain this point more clearly in the text and describe the data - e.g. As expected, depletion of INTS11 leads to increased POINT-seq signal at the 3' end of snRNAs, consistent with defects in transcriptional termination. This is highlighted by the RNU5A-1 and RNU5B-1 loci (Figure 2B).

We agree and have added more context to clarify this.

I would suggest adjusting the scale of the heatmap in Figure 2E - I think it would be easier to interpret if the value of 0 was white - with >0 a gradient of orange and <0 a gradient of blue (as is done in Figure 1C). I think making this change would make the point as written in the text clearer i.e. "heatmap analysis demonstrates the dominant impact of INTS11 on antisense versus sense transcription at most promoters (Figure 2E)." I'm assuming most of the sense transcription would be white (more clearly unchanging) when the scale is adjusted.

We agree and have done this. The reviewer is correct that most sense transcription is unchanged by INTS11 loss. However, as we alluded to in the original submission, a subset of transcripts shows a promoter-proximal increase after INTS11 depletion. We have expanded the analyses of this effect (see responses to other comments) but stress that it is neither as general nor as large as the antisense effect.

The authors make the point that there is mildly increased transcription over the 5' end of some genes upon INST11 depletion and show a track (Supplementary Fig 2A). It is not immediately obvious from the presentation of the meta-analysis in Figure 2D how generalisable this statement is. Perhaps the size of the panel or thickness of the lines in Figure 2D could be adjusted so that the peak of the control (in blue) could be seen. Perhaps an arrow indicating the peak could be added? I'm assuming the peak at the TSS is slightly lower in the control compared to INST11 depletion based on the authors' statement.

We have provided multiple new analyses of this data to highlight where there are promoter-proximal effects of INTS11 loss in the sense direction. Please see our response to the public review of reviewer 1 and new supplemental figures 2A, 2B, 4A and 4B which highlight the sense transcription increased in the absence of INTS11.

The authors label Figure 4 "Promoters lose their directionality when CDK9 is inhibited" - but in INST11 depleted cells treated with CDK9i they find that there still is a bias towards sense transcription. Suggested edit "Some promoter directionality is lost when CDK9 is inhibited" or similar.

We agree and have made this change.

The authors conclude that INTS11-mediated effects are the result of perturbation of the catalytic activities of Integrator, the authors should perform rescue experiments with the catalytically dead E203Q-INTS11 mutant.

This is a very good suggestion and something we had intended to pursue. However, as we will describe below (and shown in Supplemental Figure 4G), there were confounding issues with this experiment.

The E203Q mutant of INTS11 is widely used in the literature to test for catalytic functions of INTS11. However, we have found that this mutation impairs the ability of INTS11 to bind other Integrator modules in cells. Based on co-immunoprecipitation of flag-tagged WT and E203Q derivatives, INTS1 (backbone module), 10 (tail module), and 8 (phosphatase module) all show reduced binding to E203Q vs. WT. Because E203Q INTS11 is defective in forming Integrator complexes, rescue experiments might not fully distinguish the effects of INTS11 activity from those caused by defects in complex assembly. While this may at first seem unexpected, in the analogous 3’ end processing complex, catalytic mutants of CPSF73 (which is highly related to INTS11) negatively affect its interaction with other complex members (Kolev and Steitz, EMBO Reports 2005).

We hypothesise that INTS11 activity is most likely involved in attenuating promoter-proximal transcription, but we cannot formally rule out other explanations and discuss this in our revision. Regardless of how INTS11 attenuates transcription, our main conclusion is on its requirement to terminate antisense transcription whether this involves its cleavage activity or not.

The authors suggest that CDK9 modulates INTS11 activity/assembly and suggest this may be related to SPT5. Is there an effect of CDK9 inhibition on the snRNA's highlighted in Figure 2B?

We believe that snRNAs are different from protein-coding genes concerning CDK9 function. Shona Murphy’s lab previously showed that, unlike protein-coding genes, snRNA transcription is insensitive to CDK9 inhibition, and that snRNA processing is impaired by CDK9 inhibition (Medlin et al., EMBO 2003 and EMBO 2005). We reproduce these findings by metaanalysis of 15 highly expressed and well-separated snRNAs and by qRT-PCR of unprocessed RNU1-1, RNU5A-1 and RNU7-1 snRNA following CDK9 inhibition. We observe snRNA read-through by POINT-seq following INTS11 loss whether CDK9 is inhibited or not (left panel, below). Note the higher TES proximal signal in CDK9i conditions, which likely reflects the accumulation of unprocessed snRNA as validated by qPCR for three example snRNAs (right panel, below).

For Figure 4, would similar results be observed using inhibitors targeting other transcriptional CDKs such as CDK7,12/13?

In response to this suggestion, we analysed four selected protein-coding transcripts (the same 4 that we used to validate the CDK9i results) by qRT-PCR in a background of CDK7 inhibition using the THZ2 compound (new Supplemental Figure 4E). THZ2 suppresses transcription from these genes as expected. Interestingly, expression is restored by co-depleting Integrator, recapitulating our findings with CDK9 inhibition. As CDK7 is the CDK-activating kinase for CDK9, its inhibition will also inhibit CDK9 so THZ2 may simply hit this pathway upstream of where CDK9 inhibitors. Second, CDK7 may independently shield transcription from INTS11. We allude to both interesting possibilities.

What happens to the phosphorylation state of anti-sense engaged RNAPII when INTS11 is acutely depleted and/or CDK9 is inhibited? This could be measured by including Ser5 and Ser2 antibodies in the sPOINT-seq assay and complemented with Western Blot analysis.

We have performed the western blot for Ser5 and Ser2 phosphorylation as suggested. Both signals are mildly enhanced by INTS11 loss, which is consistent with generally increased transcription. Ser2p is strongly reduced by CDK9 inhibition, which is consistent with the loss of nascent transcription in this condition. Interestingly, both modifications are partly recovered when INTS11 is depleted in conjunction with CDK9 inhibition. This is consistent with the effects that we see on POINT-seq and shows that the recovered transcription is associated with some phosphorylation of RNAPII CTD. This presumably reflects the action(s) of kinases that can act redundantly with CDK9.

We have not performed POINT-seq with Ser5p and Ser2p antibodies under these various conditions. Our rationale is that our existing data uses an antibody that captures all RNAPII (regardless of its phosphorylation status), which we feel most comprehensively assays transcription in either direction. Moreover, the lab of Fei Chen (Hu et al., Mol Cell 2023) recently published Ser5p and Ser2p ChIP-seq following INTS11 loss. By ChIP-seq, they observe a bigger increase in antisense RNAPII occupancy vs. sense providing independent and orthogonal support for our POINT-seq data. Interestingly, this antisense increase is not paralleled by proportional increases in Ser5p or Ser2p signals. This suggests that the unattenuated antisense transcription resulting from INTS11 loss does not have high Ser5p or Ser2p. Since CDK7 and 9 are major Ser5 and 2 kinases, this supports our model that their activity is less prevalent for antisense transcription. We now discuss these data in our revision.

The HIV reporter RNA experiments should be performed with the CDK9 inhibitor added to the experimental conditions. Presumably CDK9 inhibition would result in no upregulation of the reporter upon addition of TAT and/or dTAG. Perhaps the amount of TAT should be reduced to still have a dynamic window in which changes can be detected. It is possible that reporter activation is simply at a maximum. Can anti-sense transcription be measured from the reporter?

We have performed the requested CDK9 inhibitor experiment to confirm that TAT-activated transcription from the HIV promoter is CDK9-dependent (new supplemental figure 4F). Consistent with previous literature on HIV transcription, CDK9 inhibition attenuates TAT-activated transcription. Importantly, and in line with our other experiments, depletion of INTS11 results in significant restoration of transcription from the HIV promoter when CDK9 is inhibited. Thus, TAT-activated transcription is CDK9-dependent and, as for endogenous genes, CDK9 prevents attenuation by INTS11.

While TAT-activated transcription is high, we do not think that the plasmid is saturated. When considering this question, we revisited previous experiments using this system to study RNA processing (Dye et al., Mol Cell 1999, Cell 2001, Mol Cell 2006). In these cases, mutations in splice sites or polyadenylation sites have a strong effect on RNA processing and transcription around HIV reporter plasmids. Effects on transcription and RNA processing are; therefore, apparent in the appropriate context. In contrast, we find that the complete elimination of INTS11 has no impact on RNA output from the HIV reporter. Our original experiment assessing the impact of INTS11 loss in +TAT conditions used total RNA. One possibility is that this allows non-nascent RNA to accumulate which might confound our interpretation of INTS11 effects on ongoing transcription. However, the new experiment described in the paragraph above was performed on chromatin-associated (nascent) RNA to rule this out. This again shows no impact of INTS11 loss on HIV promoter-derived transcription in the presence of TAT.

To our knowledge, antisense transcription is not routinely assayed from plasmids. They generally employ very strong promoters (e.g. CMV, HIV) to drive sense transcription. Crucially, their circular nature means that RNAPII going around the plasmid could interfere with antisense transcription coming the other way which does not happen in a linear genomic context. This is why we restricted our use of plasmids to looking at the effects of stimulated CDK9 recruitment (via TAT) on transcription rather than promoter directionality.

The authors should clearly state how many replicates were performed for the genomics experiments. Ideally, a signal should be quantified and compared statistically rather than relying on average profiles only.

We have stated the replicate numbers for sequencing experiments in the relevant figure legends. All sequencing experiments were performed in at least two biological replicates, but often three. In addition, we validated their key conclusions by qPCR or with orthogonal sequencing approaches.

Reviewer #3 (Recommendations For The Authors):

The authors provide strong evidence in support of their claims.

ChIP-seq of pol2S5 and S2 upon INST11 and CDK9 inhibition will strengthen the observation that transcription in the sense direction is more efficient.

We view the analysis of total RNAPII as the most unbiased way of establishing how much RNAPII is going one way or the other. Importantly, ChIP-seq was very recently performed for Ser2p and Ser5p RNAPII derivatives in the lab of Fei Chen (Hu et al., Mol Cell 2023). Their data shows that loss of INTS11 increases the occupancy of total RNAPII in the antisense direction more than in the sense direction, which is consistent with our finding. Interestingly, the increased antisense RNAPII was not paralleled with an increase in Ser2p or Ser5p. This suggests that, following INTS11 loss, the unattenuated antisense transcription is not associated with full/normal Ser2p or Ser5p. These modifications are normally established by CDK7 and 9; therefore, this published ChIP-seq suggests that they are not fully active on antisense transcription when INTS11 is lost. This supports our overall model that CDK9 (and potentially CDK7 as suggested for a small number of genes in new Supplemental Figure 4E) is more active in the sense direction to prevent INTS11-dependent attenuation. We now discuss these data in our revision.

In Supplementary Figure 2, the eRNA expression increases upon INST11 degradation, I wonder if the effects of this will be appreciated on cognate promoters? Can the authors test some enhancer:promoter pairs?

We noticed that some genes (e.g. MYC) that are regulated by enhancers show reduced transcription in the absence of INTS11. Whilst this could suggest a correlation, the transcription of other genes (e.g. ACTB and GAPDH) is also reduced by INTS11 loss although they are not regulated by enhancers. A detailed and extensive analysis would be required to establish any link between INTS11-regulated enhancer transcription and the transcription of genes from their cognate promoters. We agree that this would be interesting, but it seems beyond the scope of our short report on promoter directionality.

Line 111, meta plot was done of 1316 genes. Details on this number should be provided. Overall, the details of methods and analysis need improvement. The layout of panels and labelling on graphs can be improved.

We have now explained the 1316 gene set. In essence, these are the genes separated from an expressed neighbour by at least 10kb. This distance was selected because depletion of RBBP6 induces extensive read-through transcription beyond the polyadenylation site of protein-coding genes. To avoid including genes affected by transcriptional read-through from nearby transcription units we selected those with a 10kb gap between them. This was the only selection criteria so is unlikely to induce any unintended biases. Finally, we have added more information to the figure panels and their legends, which we hope will make our manuscript more accessible.

Version 2

Abstract

Abstract

RNA polymerase II (RNAPII) transcription initiates bidirectionally at many human protein-coding genes. Sense transcription usually dominates and leads to messenger RNA production, whereas antisense transcription rapidly terminates. The basis for this directionality is not fully understood. Here, we show that sense transcriptional initiation is more efficient than in the antisense direction, which establishes initial directionality. After transcription begins, the opposing functions of Integrator (INTS11) and cyclin-dependent kinase 9 (CDK9) maintain directionality. Specifically, INTS11 terminates antisense transcription, whereas sense transcription is protected from INTS11-dependent attenuation by CDK9 activity. Strikingly, INTS11 attenuates transcription in both directions upon CDK9 inhibition, and the engineered recruitment of CDK9 desensitises transcription to INTS11. Therefore, the preferential initiation of sense transcription and the opposing activities of CDK9 and INTS11 explain mammalian promoter directionality.

Introduction

In humans, most protein-coding genes initiate transcription bidirectionally. Usually, only sense transcription efficiently elongates and leads to mRNA synthesis. Antisense transcription terminates within a few kilobases (kb), and the short non-coding (nc)RNA is degraded1. Similar asymmetry frequently occurs in unicellular eukaryotes (e.g., budding yeast) and at some plant promoters2,3. Thus, promoter directionality is a broadly observed phenomenon. Interestingly, bidirectionality is the ground state of promoters and directionality is acquired over evolutionary time4. This is proposed to be through a combination of DNA sequences and proteins that favour the directional initiation and elongation of transcription.

The present explanation for the directionality of mammalian RNAPII promoters involves the arrangement of U1 snRNA binding sites and polyadenylation signals (PASs). U1 snRNA promotes RNAPII elongation through protein-coding genes by binding to RNA and preventing early termination5–7. RNA-bound U1 prevents early termination by inhibiting PASs and antagonising other attenuation mechanisms, which include the PP1 regulator, PNUTS, and the Restrictor complex8,9. In contrast, U1 binding sites are rarer in short antisense transcripts, which are often rich in PAS sequences that are proposed to promote transcriptional termination10. This model predicts that polyadenylation factors control a large fraction of antisense transcriptional termination.

The multi-subunit Integrator complex also regulates promoter-proximal transcription11–15. The Integrator complex comprises the backbone, arm/tail, phosphatase, and endonuclease modules11. Its endoribonuclease is INTS11, and its phosphatase activity is mediated by INTS6 and protein phosphatase 2A (PP2A). INTS11 endonuclease broadly affects promoter-proximal transcriptional attenuation whereas INTS phosphatase is proposed to regulate the escape of RNAPII into elongation16,17. INTS6/PP2A phosphatase functionally antagonises CDK9, which is vital for RNAPII promoter escape and productive elongation18. By this model, CDK9 activity promotes elongation across protein-coding genes and INTS6/PP2A opposes it.

We tested the prediction that PAS factors control transcription directionality by terminating antisense transcription, but this is not usually the case. Instead, we find that promoter directionality is often conferred by preferential initiation in the sense direction and is thereafter maintained by INTS11 and CDK9. The termination of antisense transcription is constitutively INTS11-dependent, whereas sense transcription is hypersensitive to INTS11 only when CDK9 activity is simultaneously inhibited. We hypothesise that CDK9 activity protects sense transcription from attenuation by INTS11 and that reduced CDK9 activity in the antisense direction exposes RNAPII to INTS11-dependent termination.

Results

Polyadenylation factor depletion does not increase or extend antisense transcription

The current explanation of mammalian promoter directionality invokes early PAS-dependent termination of antisense transcription10. This is partly based on the direct detection of polyadenylated antisense transcripts, which provides evidence that some of their transcriptional termination is PAS-dependent. However, there are non-polyadenylated antisense RNAs that might be attenuated in other ways19. To test the contribution of PAS-dependent termination toward antisense transcriptional termination and promoter directionality we tagged RBBP6 with the dTAG degron. Because RBBP6 is required to activate the PAS cleavage machinery20,21, its depletion should inhibit any PAS-dependent transcriptional termination. Three homozygous dTAG-RBBP6 clones were isolated (Supplemental Figure 1A) and tagged RBBP6 was efficiently depleted after exposure to the dTAGv-1 degrader (Figure 1A). To test the contribution of RBBP6 to nascent transcription, we used POINT (Polymerase Intact Nascent Transcript)-seq22, which maps full-length RNA extracted from immunoprecipitated RNAPII.

RBBP6 loss disrupts PAS-dependent termination of sense transcription.

A. Western blot demonstrating the depletion of dTAG-RBBP6 over a time course of dTAGv-1 addition. SPT5 serves as a loading control.

B. Genome browser track of NEDD1 in POINT-seq data from dTAG-RBBP6 cells treated or not (2hr) with dTAGv-1. RBBP6 depletion induces a termination defect in the protein-coding direction (downstream of the indicated PAS) but not the upstream antisense direction. The y-axis shows Reads Per Kilobase per Million mapped reads (RPKM).

C. Metaplot of POINT-seq data from dTAG-RBBP6 cells treated (RBBP6 dep) or not (Ctrl) (2hr) with dTAGv-1. This shows 1316 protein-coding genes selected as separated from any expressed transcription unit by ≥10kb. Signals above and below the x-axis are sense and antisense reads, respectively. The y-axis scale is RPKM. TSS=transcription start site; TES=transcription end site (this marks the PAS position). This is an average of two biological replicates.

D. Heatmap representation of the data in C, which displays signal as a log2 fold change (log2FC) in RBBP6 depleted versus un-depleted conditions. This is an average of two biological replicates.

Figure 1B shows POINT-seq coverage over NEDD1 including the upstream antisense transcript. RBBP6 loss causes a termination defect beyond the PAS shown by the extended POINT-seq signal beyond the NEDD1 gene. However, RBBP6 does not affect the termination of upstream antisense RNA which is not extended when RBBP6 is depleted. Because RBBP6 loss induces strong termination defects beyond the PAS, we generalised its effects using genes separated from their neighbours by ≥10kb. Metaplot and heatmap analyses across these 1316 genes confirm that RBBP6 loss causes a general termination defect downstream of protein-coding genes but has little impact on antisense transcription (Figures 1C and D). Although many antisense transcripts contain multiple AAUAAA sequences, most still terminate RBBP6-independently (Supplementary Figures 1B and C). Similarly, we recently showed that the PAS-dependent 5’◊3’ exonucleolytic torpedo terminator, XRN2, does not affect antisense transcriptional termination9. Therefore, although some antisense transcripts are polyadenylated, most antisense transcription can terminate using PAS-independent mechanisms. Consequently, PAS-dependent termination cannot fully explain the promoter-proximal attenuation of antisense transcription.

Integrator depletion increases and extends antisense transcription

Our data argue that PAS-independent termination mechanisms control a large fraction of antisense transcription. A major PAS-independent termination pathway is driven by the Integrator complex, which was first identified as the 3’ end processing complex for snRNAs23.

Several reports show that Integrator terminates transcription from most promoters, including those that initiate antisense transcription, so it might affect promoter directionality11,16,17. To analyse this, we tagged its endonucleolytic subunit (INTS11) with a dTAG degron24, which enables rapid depletion of its catalytic activity (Figure 2A). We then performed POINT-seq on INTS11-dTAG cells depleted or not of INTS11 to assay the global impact of INTS11 on RNAPII transcription. Integrator is established to control the transcription of snRNAs, and this was detected in our POINT-seq, which demonstrates a clear extension of read density at RNU5A-1 and RNU5B-1 loci (Figure 2B).

INTS11 loss disrupts the termination of antisense transcription.

A. Western blot demonstrating homozygous tagging of INTS11 with dTAG and the depletion of INTS11-dTAG after 1.5hr treatment with dTAGv-1. INTS8 serves as a loading control.

B. Genome browser track showing POINT-seq signal over RNU5A-1 and RNU5B-1 in INTS11-dTAG cells treated or not (1.5hr) with dTAGv-1. Note, that INTS11 depletion induces a termination defect in each case. Y-axis shows RPKM following spike-in normalisation.

C. Genome browser track showing POINT-seq signal over PGK1 and its upstream antisense region in INTS11-dTAG cells treated or not (1.5hr) with dTAGv-1. Y-axis shows RPKM following spike-in normalisation.

D. Metaplot of POINT-seq data from INTS11-dTAG cells treated or not (1.5hr) with dTAGv-1. This shows the same 1316 genes used in Figure 1C. Signals above and below the x-axis are sense and antisense reads, respectively. Y-axis shows RPKM following spike-in normalisation. This is an average of three biological replicates.

E. Heatmap representation of the data in D, which displays signal as a log2 fold change (log2FC) in INTS11 depleted versus undepleted conditions over a region 3kb upstream and downstream of annotated TSSs. This is an average of three biological replicates.

As exemplified by PGK1 (Figure 2C), INTS11 loss does not affect the termination of protein-coding transcription but causes a strong upregulation of antisense transcription. Meta-analysis of the same gene set used for RBBP6 POINT-seq shows the generality of antisense transcriptional attenuation via INTS11 (Figure 2D). A heatmap analysis of transcription 3kb upstream and downstream of the same transcription start sites (TSSs) demonstrates the dominant impact of INTS11 on antisense vs. sense transcription at most of these promoters (Figure 2E). Although the transcription over protein-coding genes is less affected by INTS11 loss, an increase in promoter-proximal transcription is apparent in some cases (Supplemental Figure 2A). The most strongly affected genes are lowly expressed, which is consistent with recent findings (Supplemental Figure 2B)12–17. Overall, INTS11 frequently attenuates antisense transcription whereas a smaller fraction of sense transcription is affected. RNAPII.

The hypersensitivity of antisense transcripts to INTS11 might be due to the makeup of antisense RNAPII or some other promoter feature. To interrogate this further, we analysed two other promoter classes: those where protein-coding transcripts are initiated in both directions and those that initiate the bidirectional transcription of unstable enhancer (e)RNAs. In the former case, both transcripts are extended and stable like most sense transcripts at directional promoters. In the latter case, both are short and unstable like most antisense transcripts at directional promoters. When both directions are protein-coding, INTS11 depletion causes modest reductions in transcription (Supplemental Figure 2C). Conversely, eRNA transcription is bidirectionally upregulated upon INTS11 elimination like the antisense transcripts in (Supplemental Figure 2D). We conclude that short ncRNAs are more strongly affected by INTS11 than protein-coding transcripts. At directional promoters this results in the attenuation of antisense transcription.

Transcription initiates more efficiently in the sense direction

The increased antisense POINT-seq signal following INTS11 loss is consistent with defects in transcriptional termination. However, it could result from increased transcriptional initiation. We were also interested in whether preferential sense initiation could contribute to mammalian promoter directionality as was proposed in budding yeast4. To precisely resolve directional aspects of initiation and assay any impact of INTS11, we devised a variant of POINT-seq called short (s)POINT. Briefly, sPOINT follows the POINT-seq protocol, but library preparation employs the selective amplification of 5’ capped RNAs <150nts (Figure 3A, Supplemental Figure 3A, Experimental Procedures). The sPOINT signal over PGK1 exemplifies this and demonstrates a very restricted signal close to the TSS (Figure 3B). Figure 3C compares the meta profile of POINT- and sPOINT-seq on 684 well-expressed and well-spaced (≥10kb from neighbours) genes to highlight the full read coverage obtained by POINT-seq and the tightly restricted, TSS-proximal, sPOINT signal. Because most capped RNAPII-associated RNA <150nts are promoter-proximal, sPOINT effectively assays the promoter-proximal RNAPII pause.

Sense direction transcription initiation is more efficient and focused.

A. Schematic of sPOINT-seq protocol. The POINT-seq protocol is followed, in which chromatin is isolated and engaged RNAPII is immunoprecipitated. Short transcripts are preferentially amplified during library preparation (see Experimental Procedures for full details).

B. Comparison of POINT-(top trace) and sPOINT-seq (lower trace) on PGK1. Y-axis units are RPKM.

C. Metaplot comparison of POINT-(top plot) and sPOINT-seq (lower plot) profiles across the 684 highest expressed protein-coding that are separated from expressed transcription units by ≥10kb (top ∼50%). Signals above and below the x-axis are sense and antisense reads, respectively. Y-axis shows RPKM following spike-in normalisation.

D. Top metaplot shows full read coverage for sPOINT-seq performed in INTS11-dTAG cells treated or not (1.5hr) with dTAGv-1 at the promoters of the top expressed 20% of protein-coding genes. The lower metaplot is the same data but only the 5’ end of each read is plotted. The y-axis signals are RPKM following spike-in normalisation. Two biological replicates of sPOINT were performed.

E. Genome browser track of PGK1 promoter region in sPOINT-seq. This showcases the focused sense TSS (black arrows) and the dispersed antisense reads (brackets). Note the higher y-axis scale (RPKM) for sense vs. antisense.

F. Metaplot zoom of the antisense TSS signals deriving from the lower plot in D. This makes clear the dispersed sites of initiation. The Y-axis scale is RPKM following spike-in normalisation. Two biological replicates of sPOINT were performed.

The sPOINT-seq metaplot in Figure 3B shows a higher signal in the sense direction, suggesting that more efficient transcriptional initiation establishes promoter directionality. To assay this further and test the impact of INTS11, we performed sPOINT-seq in INTS11-dTAG cells treated or not with dTAGv-1 and plotted the coverage over the promoters of well-expressed protein-coding genes (3060 promoters, Figure 3D). As sPOINT-seq maps the 5’ and 3’ ends of these reads, precise TSSs are mapped at single-nucleotide resolution and plotted in the lower meta profile in Figure 3D. This analysis once again shows a higher sPOINT signal in the sense direction. In addition, the lower TSS mapped plot also shows that sense transcription initiated in a more focused manner compared to in the antisense direction. Figure 3E (PGK1) and supplemental figure 3B (ACTB) exemplify these features on individual genes, and the metaplot in Figure 3F highlights the dispersed nature of antisense TSSs. Quantitation of the TSS-derived sense vs. antisense signal confirms a higher read count in the sense direction in untreated cells (Supplemental Figure 3C). The most noticeable (although still modest) effect of INTS11 loss on sPOINT-seq profiles is a mild signal reduction. This shows that INTS11 depletion does not generally enhance transcriptional initiation in either direction. This slight reduction in sPOINT-seq signal could result from less transcriptional initiation or less efficient pausing if, for example, INTS11 loss allows more RNAPII to escape the promoter. Although a lower resolution technique, RNAPII ChIP-seq confirmed these promoter characteristics and the mild impact of INTS11 (Supplemental Figures 3D and E). Overall, these data show that directionality might be established by more efficient initiation in the sense direction.

CDK9 inhibition sensitises sense transcription to INTS11

After transcription initiates, further control of directionality is evident because sense transcription goes further than antisense transcription. If INTS11 regulates this, an opposing force is required to promote elongation. INTS11 affects transcription very early, occupies promoters, and becomes less active as RNAPII moves into elongation16–18,25,26. Therefore, if INTS11 is counteracted to allow sense transcription, any responsible mechanism needs to act early. One of the first transcriptional checkpoints involves the phosphorylation of RNAPII and other factors by CDK9, which releases promoter-proximally paused RNAPII into elongation27. During this process, the Integrator-associated phosphatase, PP2A, antagonizes CDK9 and presumably regulates the sensitivity of RNAPII to INTS1118. Because Integrator phosphatase has little effect on antisense transcription16, we hypothesised that INTS11 sensitivity and CDK9 activity are inversely correlated to maintain directionality after initiation.

Our hypothesis predicts that sense transcription will be attenuated via INTS11 when CDK9 is inactive. To test this genome-wide, we depleted INTS11 from INTS11-dTAG cells in the presence or absence of a specific CDK9 inhibitor (NVP-228) and performed POINT-seq. As exemplified by TARDBP, the depletion of INTS11 alone caused an antisense transcriptional termination defect with a milder impact on the protein-coding sense direction (Figure 4A). As expected, NVP-2 treatment reduced transcription over the protein-coding gene body. Antisense transcription also displays some CDK9 sensitivity. Importantly, in NVP-2-treated cells, INTS11 loss increased transcription in both directions. This contrasts with the dominant antisense effect deriving from just depleting INTS11 (refer to Figure 2). As such, CDK9 activity prevents sense transcription from being attenuated by INTS11. This is a genome-wide trend as shown in the metaplots in Figures 4B and C. A heatmap of the effect of INTS11 loss after CDK9 inhibition shows the bidirectional upregulation of transcription at the same promoters assayed in Figure 2 (Figure 4D). Although this experiment employed longer INTS11 depletion than that in Figure 2 (2.5hr vs. 1.5hr to allow concurrent CDK9 inhibition), its antisense effects remain general, and the affected protein-coding genes strongly overlapped (Supplemental Figure 4A-C). Lastly, when CDK9 and INTS11 are both compromised, the POINT-seq signal remains higher in the sense vs. antisense direction (black line, Figure 4C). This is consistent with our sPOINT-based finding that initiation of sense transcription is generally more efficient.

Some promoter directionality is lost when CDK9 is inhibited.

A. Genome browser track of TARDBP in POINT-seq data derived from INTS11-dTAG cells either untreated, dTAG treated, NVP-2 treated or dTAG and NVP-2 treated (2.5 hr). Signals above and below the Y-axis are sense and antisense reads, respectively. The Y-axis scale shows RPKM following spike-in normalisation.

B. Metaplot of POINT-seq in INTS11-dTAG cells depleted or not of INTS11 and treated or not with NVP-2 (2.5 hr). This uses the same gene set as Figure 1C. The regions 3kb upstream and downstream of genes are included. Y-axis units are RPKM following spike-in normalisation. This is an average of two biological replicates.

C. Metaplot of the same CDK9i + and -dTAG data shown in B but zoomed into the region 3kb upstream and downstream of the TSS. This is an average of three biological replicates.

D. Heatmap representation of the data in C, which displays signal as a log2 fold change (log2FC) in INTS11 depleted versus un-depleted conditions covering a region 3kb upstream and downstream of the TSS. This is an average of three biological replicates.

E. Western blot for total RNAPII and RNAPII phosphorylated on Ser2/5 (Ser2/5p) in INTS11-dTAG cells treated or not with dTAGv-1 and/or NVP-2 (as indicated). dTAG-mediated INTS11 depletion is shown in the anti-HA blot and SPT5 us used as a loading control.

F. qRT-PCR analysis of INTS11-dTAG cells transfected with the HIV reporter construct with or without TAT then depleted or not of INTS11. Quantitation shows signals relative to those obtained in the presence of INTS11 and the absence of TAT after normalising to MALAT1 RNA. n=3. Error bars show standard deviation. ** denoted p=0.01. Note that INTS11 depletion was performed concurrently with transection (14hr in total).

G. Model for promoter directionality depicting higher levels of focused transcriptional initiation in the sense direction together with opposing gradients of CDK9 and INTS11 activity that peak in sense and antisense directions, respectively.

Following CDK9 inhibition and INTS11 loss, the largest recovery of protein-coding transcription is often over the 5’ end of genes. This presumably reflects poor elongation without CDK9 activity even when INTS11 is absent. RNAPII elongation is associated with phosphorylation of the C-terminal domain (CTD) of its largest subunit, RPB1, and most frequently occurs on Serine 5 or 2 (Ser5/2p) of its heptad repeat29,30. We assayed the effects of CDK9 inhibition and INTS11 depletion on these two modifications by western blotting (Figure 4E). As previously shown28, CDK9 inhibition by NVP-2 suppresses Ser2p. However, both Ser5p and Ser2p are enhanced by INTS11 loss whether CDK9 is active or not. Thus, unattenuated transcription after INTS11 depletion is associated with some CTD phosphorylation. Where NVP-2 is employed, kinases redundant with CDK9 presumably phosphorylate RNAPII.

We confirmed the CDK9-mediated suppression of INTS11 on four selected protein-coding genes using an alternative inhibitor, 5, 6-dichloro-1-β-D-ribofuranosylbenzimidazole (DRB) (Supplemental Figure 4D). Interestingly, inhibition of CDK7 with THZ2 sensitised the same selected protein-coding transcripts to INTS11 attenuation (Supplemental Figure 4E). This may be because CDK7 is required to activate CDK931, or because Integrator additionally targets RNAPII when CDK7 is inactive. This result reinforces the idea that the successful execution of early transcriptional checkpoints may overcome Integrator-mediated attenuation.

If CDK9 counteracts INTS11, its recruitment should prevent transcriptional attenuation by Integrator. To assay this, we employed a plasmid where transcription is driven by the human immunodeficiency virus (HIV) promoter. Transcription from the HIV promoter results in synthesis of the trans-activating response (TAR) element and promoter-proximal RNAPII pausing. Pause release requires the trans-activator of transcription (TAT), which promotes RNAPII elongation by recruiting CDK932. INTS11 suppresses transcription from the HIV promoter when TAT is absent33. To assay whether CDK9 affects this process, INTS11-dTAG cells were transfected with the HIV reporter with or without TAT before treatment or not with dTAGv-1. Transcription from the reporter was then analysed by qRT-PCR (Figure 4F). INTS11 loss induces HIV transcription in the absence of TAT. However, TAT strongly stimulates transcription (∼200 fold) and desensitises it to INTS11 loss (slightly less reporter RNA was recovered). Therefore, TAT-mediated CDK9 recruitment alleviates INTS11-dependent attenuation of transcription. In line with endogenous protein-coding genes, CDK9 inhibition sensitises TAT-activated transcription to INTS11 (Supplemental Figure 4F). Overall, our data strongly suggest that CDK9 activity counteracts transcriptional attenuation by INTS11.

Discussion

Most studies on mammalian promoter directionality have focused on understanding how antisense transcription terminates and how RNAPII elongation is enabled in the sense direction. However, we show that preferential sense transcriptional initiation significantly contributes toward the directionality of many mammalian promoters. More analyses are required to elucidate the mechanism, but we hypothesise that a unidirectional arrangement of at least some core promoter elements favours transcription in the protein-coding direction. Consistently, the evolution of promoter elements explains how a bidirectional promoter “ground state” acquired directionality in yeast4. Consistent with our finding, focused transcription initiation is associated with highly expressed mammalian genes and promoters with clearly defined elements (e.g. TATA)34. Thus, the lower level and dispersed nature of antisense initiation may reflect the suboptimal orientation of promoter elements or the opportunistic transcription of open chromatin in promoter regions.

We hypothesise that INTS11 controls transcriptional attenuation via its endonuclease activity. An inactive mutant of INTS11 (E203Q) is widely used to interrogate its catalytic function23; however, we discovered that it poorly associates with other Integrator components compared to wild-type INTS11 (Supplemental Figure 4G). Therefore, when employed in cells, it may not isolate the effects of INTS11 activity from those requiring an intact Integrator complex. Similarly, catalytic mutations in the highly related PAS endonuclease, CPSF73, disrupt its association with other cleavage and polyadenylation components35. In light of this, it is formally possible that non-catalytic consequences of INTS11 loss explain the upregulation of antisense transcription.

How CDK9 activity opposes INTS11 is unresolved, but INTS11 and SPT5 are adjacent in the RNAPII: Integrator structure36. As SPT5 is a prominent substrate of CDK937–39, its phosphorylation might evict INTS11 or prevent its association with the complex. Consistently, Integrator is enriched on promoters whereas phosphorylated SPT5 (SPT5p) is most prevalent during elongation over protein-coding gene bodies25,38. Furthermore, although antisense RNAPII occupancy is increased by INTS11 loss, published data suggest that this is not paralleled by increases in its Ser5p and Ser2p forms16. As these modifications require CDK7 and CDK9, this provides further support that these activities are less prevalent for antisense vs. sense transcription.

While INTS11 and CDK9 activities are the clearest on antisense and sense transcription, respectively, this is not binary. Some antisense transcription is NVP-2 sensitive, which indicates CDK9 activity on some RNAPII (Figure 4 and40). CDK9-sensitive antisense transcription could undergo INTS11-independent attenuation, which might account for the polyadenylated antisense RNA observed previously10,41. Consistently, some antisense transcription is targeted by the poly(A) exosome targeting connection, which operates on cleaved and polyadenylated transcripts42,43. Reciprocally, promoter-proximal transcription from lowly expressed protein-coding transcription is INTS11-sensitive, which could reflect low CDK9 activity at those loci. We and others recently described the Restrictor complex as restraining antisense transcription9,44,45. Unlike Integrator, which associates with NELF-bound RNAPII36, Restrictor is proximal to the distinct PAF-bound RNAPII9. Thus, various complexes may target different forms of RNAPII before and after CDK9 activity. Even so, our data suggests that Integrator influences a greater volume of antisense RNAPII than the cleavage and polyadenylation pathway.

In addition to CDK9, multiple studies demonstrate a role for U1 snRNA in promoting elongation through protein-coding genes5,6,8,46. U1 promotes RNAPII elongation and shields transcription from early PAS-dependent termination and the Restrictor complex5,9. The U1-mediated suppression of premature PAS-dependent termination inspired the original model of promoter directionality. As U1 sites are rarer in antisense RNAs, their termination was hypothesised to be driven by early PASs that would consequently be active. Our RBBP6 data (and other published data on PAS termination factors9) argues that a large fraction of antisense transcriptional termination is PAS-independent. Nevertheless, the lack of U1 sites may limit RNAPII elongation and render it sensitive to Integrator (or Restrictor). Accordingly, it will be interesting to test whether U1 regulates the sensitivity of transcription to Integrator.

In conclusion, we provide a new model for mammalian promoter directionality and the subsequent control of bidirectional transcription (Figure 4G). sPOINT demonstrates more efficient sense transcription initiation, which establishes directionality. Early events dictate the decision to elongate or attenuate transcription. In the sense direction, CDK9 activity prevents attenuation by Integrator to favour productive elongation, whereas antisense transcription is hypersensitive to INTS11 and terminates early. As INTS11 occupies most promoters and CDK9 activity is near-universally required to achieve protein-coding transcription, this model can generally explain how directionality is initiated and maintained. Our new sPOINT-seq approach will be valuable in elucidating additional aspects of promoter-proximal transcriptional regulation.

Supporting information